Correlation tests

### 2025-03-27 analysis/manuscript.tex
Proofreading

hoaddata, developed and maintained by the author to support open access
monitoring and research (Jahn, 2025), is an R data package that
regularly collects information on hybrid open access uptake from
multiple openly available data sources (Jahn, 2024). It combines
article-level metadata from Crossref (Hendricks et al., 2020) and
OpenAlex (Priem et al., 2022) with transformative agreement information
from Transformative Agreement Data dump used by the cOAlition S Journal
Checker Tool (JCT)\footnote{\url{https://journalcheckertool.org/transformative-agreements/}},
which links journal and institutional data to agreements in the ESAC
registry.

As discussed elsewhere (Jahn, 2025), hoaddata uses a unified dataset of
various safeguarded JCT snapshots \footnote{\url{https://github.com/njahn82/jct_data}}.
Journal data were enriched with ISSN variants linked to an ISSN-L and
institution data with associated ROR-IDs representing university
hospitals or institutes of large research organisations, according to
OpenAlex' institution entity. A comprehensive exclusion of fully open
access journals using multiple journal lists (DOAJ, OpenAlex, Bielefeld)
was performed to identify hybrid journals. hoaddata relies on Crossref
for obtaining journal publication volume and open access status through
Creative Commons licence information relative to the published version
(``version of record''). Because of limited affiliation metadata in
Crossref (Eck \& Waltman, 2022), hoaddata sources affiliation metadata
from OpenAlex.

@@

Clarivate Analytics' Web of Science (WoS) is a well-established
proprietary bibliometric database consisting of several collections
(Birkle et al., 2020). The collections considered in this study were the
Science Citation Index Expanded (SCIE), the Social Sciences Citation
Index (SSCI) and the Arts \& Humanities Citation Index (AHCI).

These collections provides important data points for analysing open
access: author affiliations and roles, differentiation of journal
articles into document types representing different types of journal
contributions, such as original articles or reviews, and open access
status information derived from OurResearch's Unpaywall, the same
provider as OpenAlex. However, the Web of Science lacks information
about journals and articles under transformative agreements.

@@

Elsevier's Scopus, launched in 2004, is another widely used proprietary
bibliometric database for measuring research (Baas et al., 2020).
Similar to the Web of Science, Scopus is selective with regard to the
journals it indexes. However, its coverage is substantially more
extensive than that of the Web of Science collections considered in this
study (Singh et al., 2021; Visser et al., 2021). With detailed metadata
about article types, open access status information derived from
Unpaywall, author roles, and disambiguated affiliations, Scopus also
contains important data to assess open access uptake, although no direct
information regarding transformative agreements was available at the
time of the study.

@@

Using an enriched and unified list of hybrid journals in transformative
agreements according to the JCT, article metadata were retrieved from
each database using all ISSN variants linked to an ISSN-L.

The article metadata included DOIs, publication dates, open access
information as well as author roles and institutional affiliations.
Publication years were determined using the earliest known date of
publication in a journal. In hoaddata, this corresponded to Crossref's
issued date. For Web of Science and Scopus, the earliest publication
date was used where available, with Scopus dates specifically determined
by the KB through version tracking of the raw data.

Many transformative agreements typically cover only certain types of
journal articles, in particular original research articles including
reviews. Because of limited information on these document types in open
scholarly data (Haupka et al., 2024), hoaddata used an extended version
of Unpaywall's paratext recognition approach to exclude non-scholarly
content (Piwowar et al., 2018). To exclude conference supplements, which
are also often not covered by transformative agreements, only articles
published in regular issues, indicated by numerical pagination, were
considered. For Web of Science and Scopus, their established document
type classifications were used to identify original research articles
and reviews, referred to as original articles throughout this study.

@@

Articles in hybrid journals were considered open access when they were
made freely available under a Creative Commons license on publishers'
platforms. While hoaddata sourced this information from Crossref license
metadata, Web of Science and Scopus relied on Unpaywall as evidence
source. Unpaywall supplements Crossref metadata by parsing publisher
websites directly, addressing cases where publishers do not provide
machine-readable Creative Commons license information (Piwowar et al.,
2018). This additional parsing remains necessary despite transformative
agreement workflows recommend the depositing license information during
DOI registration (Geschuhn \& Stone, 2017). Both Web of Science and
Scopus defined hybrid open access consistently as content available
under Creative Commons licenses on publisher platforms, distinguishing
it from bronze open access that lack such explicit license information,
or use publisher-specific licenses.

@@

Author affiliations were retrieved for both first and, if available,
corresponding authors to prepare the linking between articles and
institutions covered by transformative agreements. To handle different
address variants, database-specific affiliation identifiers were used:
ROR-IDs from OpenAlex for hoaddata, affiliation enhanced names for Web
of Science, and Scopus Affiliation Identifiers for Scopus. Additionally,
ISO country codes were retrieved for each author's address to compile
country-level statistics. These country codes for Web of Science and
Scopus were provided by the KB.

@@

Based on these matching tables, articles eligible under transformative
agreements could also be retrieved from Web of Science and Scopus,
although they did not contain the ROR IDs used by the JCT. The
estimation of eligible articles followed Jahn (2025) and included a
matching of both journals and participating institutions according to
the Transformative Agreement Data dump for each of the data sources
examined. The matching also took into account the duration of agreements
according to the ESAC registry, with only those matches where an
agreement was actually in place being considered for subsequent
analysis.

@@

\centering
\begin{tabular}[t]{lrrr}
\toprule
\textbf{} & \textbf{HOAD} & \textbf{Web of Science} & \textbf{Scopus}\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Hybrid journal coverage}}\\
\hspace{1em}Active journals & 12,890 & 8,655 & 11,888\\
\hspace{1em}Active journals (core) & 12,888 & 8,655 & 11,878\\
\hspace{1em}Active journals (core) with OA & 11,348 & 8,392 & 11,313\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Publication volume}}\\
\hspace{1em}Total published articles & 9,740,015 & 8,616,053 & 8,117,644\\
\hspace{1em}Original articles & 8,158,425 & 6,708,083 & 7,317,703\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Digital Object Identifier (DOI) coverage}}\\
\hspace{1em}Articles with DOI & 9,740,015 & 7,713,796 & 8,105,112\\
\hspace{1em}Original articles with DOI & 8,158,425 & 6,695,661 & 7,314,327\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Open Access (OA) metrics}}\\
\hspace{1em}OA articles & 998,699 & 1,112,758 & 974,099\\
\hspace{1em}Original OA articles & 969,817 & 1,019,784 & 922,578\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Original articles with affiliation data}}\\
\hspace{1em}First author articles & 7,242,542 & 6,294,855 & 7,232,017\\
\hspace{1em}Corresponding author articles & 5,534,207 & 6,291,441 & 6,898,487\\
\bottomrule
\end{tabular}

@@

While hoaddata only covers articles with a DOI, Scopus and Web of
Science publication indicators were calculated using their database
identifier. A subsequent comparison of DOI coverage shows that
non-original articles in Web of Science often lacked a DOI. This was
particularly the case for meeting abstracts, which are notably prevalent
in Health Sciences journals (Melero-Fuentes et al., 2025) and are not
indexed by Scopus (Donner, 2017). Open access indicators were aggregated
by DOI, as Unpaywall only collects information on open access status for
articles with a DOI. A closer look at original articles with affiliation
data reveals a lack of data on corresponding authors in the case of
OpenAlex, the affiliation data source used by hoaddata, compared to Web
of Science and Scopus. Only about two-thirds of the articles examined
provided corresponding author affiliations. For first authors, the
proportion was 89\%. Therefore, only first author data for hoaddata were
considered in the following analysis.

@@

Journal coverage analysis revealed that 66\% (n = 7,970) of hybrid
journals included in transformative agreements were indexed in all three
databases (Figure \ref{fig-upset_coverage_results}A). The second-largest
set consisted of journals indexed in both hoaddata and Scopus,
comprising 21\% (n = 2,595) of hybrid journals. Notably, 6\% (n = 739)
of journals were exclusively contained in hoaadata, while another 6\% (n
= 748) were only found in the proprietory database Scopus. Of these, 354
were also available in the Web of Science. Upon inspection, this group
of hybrid journals exclusively covered in the proprietary data sources
mainly represented hybrid journals for which no open access evidence
could be retrieved from Crossref, the open access evidence source for
hoaddata.

@@

Using overlapping DOIs, the publication volume between 2019 and 2023 was
also calculated for each journal. Figure
\ref{fig-upset_coverage_results}C illustrates the distribution for each
combination. It shows that there is a large spread across the journals
covered by all three data sources. Furthermore, journals in this set
published more on average than journals covered by only one or two data
sources. In particular, the journals covered exclusively by hoaddata
were substantially smaller than those covered by all three sources. Upon
inspection, these were often newly launched journals, which explains the
relatively low five-year publication volume. An example is \emph{Digital
society} that published 86 articles. This hybrid journal was launched in
2022, being covered by various Springer Nature transformative agreements
since then.

@@

Figure \ref{fig-upset_coverage_results_publisher} presents the coverage
of hybrid journals in transformative agreements across data sources from
2019 to 2023 with a focus on publisher portfolios. The analysis
highlights the dominance of the three largest publishers--Elsevier,
Springer Nature, and Wiley--, which collectively accounted for 47\% of
hybrid journals and 62\% of articles published during this five-year
period. In terms of article volume, Elsevier led with 2,441,358 articles
(33\% of the total) published across 1,951 hybrid journals (16\% of the
total). Springer Nature followed with 1,247,578 articles (17\%) in
2,311, although recording the largest number of hybrid journals (19\%).
Wiley accounted for 858,939 articles (12\%) in1,382 hybrid journals
(11\%). The remaining 54 publishers collectively accounted for 2,749,847
articles (38\%) in 6,476 hybrid journals (53\%).

@@

A comparative analysis of publisher portfolios across different indexing
sets demonstrates that publisher were not represented uniformly.
Notably, Springer Nature exhibited 519 hybrid journals exclusively
indexed in both hoaddata and Scopus. This set included journals from the
Chinese Academy of Science, German-language medical journals, and
Eastern European publications including the \emph{Journal of
Mathematical Sciences}, which also publishes English-language
translations of Russian-language works. Additionally, this subset
included titles with a more broader disciplinary focus such as \emph{SN
Computer Science} and newly launched hybrid journals like \emph{Nature
Computational Science}, which started in 2021 and was indexed in Scopus
but not yet in Web of Science. The set also captured ceased journals,
providing further insights into the dynamic nature of journal portfolios
and business strategies.

@@

This section examines the uptake of open access in hybrid journals,
focusing on the influence of transformative agreements across hoaddata,
Scopus, and Web of Science. The aim was to assess whether consistent
results can be derived from these data sources despite differences in
coverage and methodologies. Following Jahn (2025), indicators were
calculated for each data source and comprise the number and proportion
of open access articles, including those enabled by transformative
agreements, from 2019 to 2023. For Web of Science and Scopus, the impact
of transformative agreements was estimated using both first and
corresponding authorships, while hoaddata indicator calculation was
limited to first-author affiliations.

@@

Figure \ref{fig-uptake_overview}A shows a moderate growth of open access
in hybrid journals, which is consistent across hoaddata (black line),
Scopus (orange line), and Web of Science (blue line). According to
hoaddata, hybrid open access uptake increased from 6.4\% (n = 85,071) in
2019 to 19\% (n = 302,358) in 2023. Similarly, Scopus recorded an growth
from 6.5\% (n = 84,648) in 2019 to 19\% (n = 322,850) in 2023. However,
Web of Science recorded higher open access uptake in early years, before
converging in 2023, from 11\% (n = 137,202) in 2019 to 20\% (n =
255,481) in 2023,suggesting an different approach towards labeling
hybrid open access by the Web of Science.

Simialry, hybrid open access by transformative agreements increased
between 2019 and 2023 (Figures \ref{fig-uptake_overview}B and C). Trends
were consistent for first (dashed line) and corresponding author (dotted
line) affiliations. According to Scopus, 479,297 open access articles
could be attributed to transformative agreements based on first author
metadata (increasing from 15,341 to 206,084) and 489,262 using
corresponding author metadata (from 15,444 to 210,816). Web of Science
recorded 493,028 open access articles via transformative agreements
using first author metadata (increasing from 21,871 to 174,126) and
500,076 using corresponding author metadata (from 22,092 to 177,030).
hoaddata, lacking corresponding author data, linked 501,649 articles to
transformative agreements but showed slower growth than Scopus and Web
of Science, from 16,010 to 183,757.

From 2021 (hoadata, Scopus) resp. 2022 (Web of Science), transformative
agreements enabled the majority of hybrid open access. For first
authors, the share ranged between 61\% (hoaddata), 64\% (Scopus), and 68
\% (Web of Science) in 2023. For corresponding authors, the shares were
slightly larger, with Scopus recording 65\% and Web of Science 69\% in
2023. However, substantial hybrid open access was still facilitated
outside transformative agreements, likely through APCs payed from
discretionary research funds (ref).

@@

When considering open access trends by publisher (see Figure
\ref{fig-uptake_publisher}), the observed differences in early uptake
rates between hoaddata and Scopus compared to Web of Science can be
largely attributed to articles published in Elsevier hybrid journals,
the largest publisher in our sample. Both hoadata and Scopus reported an
steady increase in open access uptake between 2019 and 2023 (hoaddata
from 4\% to 13\%; Scopus from 5\% to 15\%). In contrast, Elsevier's
share remained relatively constant, increasing only slightly from 14\%
to 15\% according to the Web of Science. Upon inspection, this
discrepancy primarily stemmed from articles provided under Elsevier's
user license, used for the publisher's open archive. These articles,
made freely available after an embargo period, were tagged as hybrid
open access in Web of Science, even though its documentation specified
that only articles under a CC license variant were considered. Previous
research (Haustein et al., 2024; Jahn et al., 2022) has shown that
Elsevier provided a substantial portion of its articles under this
license, explaining the relatively large and stable share of open access
over the years.

@@

Despite these differences in open access evidence, the three data
sources show consistent temporal trends in hybrid open access enabled by
transformative agreements (see Figure \ref{fig-uptake_publisher}B and
C). Wiley emerged as the fastest-growing publisher in terms of open
access uptake, with more than 30\% of articles in hybrid journals
reported as open access in 2023 across the examined data sources,
followed by Springer Nature and Elsevier. By 2023, the share of open
access enabled by transformative agreements appeared to stabilise for
all three publishers, with Wiley recording the largest share (see Figure
\ref{fig-uptake_publisher}B). Elsevier, however, showed a comparable
late uptake, consistent with the publisher's historical reluctance to
engage in negotations with library consortia. Interestingly, the
differences between first and corresponding author affiliations were
more pronounced at the publisher level. In Scopus, for example, the
share of open access via transformative agreements measured by
corresponding authorship was greater for Springer Nature in 2023 than
when using first authorship.

@@

When examining the percentage of open access articles in hybrid journals
(see Figure \ref{fig-uptake_country}B), a different pattern emerged.
Authors affiliated with institutions from medium-sized European
countries, such as Sweden, the Netherlands, Finland, and Hungary,
provided a large proportion of their articles in open access. Germany
and the United Kingdom also had approximately 40\% of their output
available as open access. In contrast, non-OECD countries showed notably
lower adoption of hybrid open access, with South Africa being the only
BRICS member well-represented in the data. The United States also
demonstrated a relatively low proportion of open access articles. These
findings were consistent across all databases. However, France was
better represented in Web of Science, likely due to its agreement with
Elsevier starting in 2019, which allowed delayed open access under the
publisher's user license (Rabesandratana, 2019). This license was not
classified as hybrid open access in Scopus or hoaddata. In all cases,
Spearman rank correlations were \(\rho > .9\), \(p < 0.001\), showing a
high level of correlations between the databases and authorship roles
considered.

Transformative agreements appeared to be a key driver of open access
growth (see Figure \ref{fig-uptake_country}C and D). OECD members
accounted for the majority of open access articles enabled by
transformative agreements. As a notable exception, South Africa also
featured prominently, as the South African National Library and
Information Consortium (SANLiC) successfully negotiated transformative
agreements with major publishers from 2022 onward.\footnote{\url{https://sanlic.ac.za/read-and-publish-agreements/}}
Results were consistent across data sources, with a strong correlation
observed. However, Wiley surplus in hoaddata led to better rankings for
countries where Wiley played a substantial role, such as Germany, where
the DEAL consortium negotiated its first transformative agreements with
Wiley that started in July 2019.

@@

@@

@@

\bibitem[\citeproctext]{ref-van_Eck_2022}
Eck, N. J. van, \& Waltman, L. (2022). \emph{Crossref as a source of
open bibliographic metadata}.
\url{https://doi.org/10.31222/osf.io/smxe5}

@@

@@

@@

>>>>

hoaddata, developed and maintained by the author to support open access
monitoring and research (Jahn, 2025), is an R data package that
regularly collects information on hybrid open access uptake from
multiple openly available data sources (Jahn, 2024). It combines
article-level metadata from Crossref (Hendricks et al., 2020) and
affilation metadata from OpenAlex (Priem et al., 2022) with
transformative agreement information from Transformative Agreement Data
dump used by the cOAlition S Journal Checker Tool (JCT)\footnote{\url{https://journalcheckertool.org/transformative-agreements/}},
which links journal and institutional data about participating research
organisations to agreements in the ESAC registry.

@@

Clarivate Analytics' Web of Science (WoS) is a well-established
proprietary bibliometric database consisting of several collections
(Birkle et al., 2020). Web of Science is a selective, base
research-focused database with dedicated panels to choose journals for
inclusion (Stahlschmidt \& Stephen, 2022; Visser et al., 2021). The
collections considered in this study were the Science Citation Index
Expanded (SCIE), the Social Sciences Citation Index (SSCI) and the Arts
\& Humanities Citation Index (AHCI).

These collections provides important data points for analysing open
access: author affiliations and roles, differentiation of journal
articles into document types representing different types of journal
contributions, such as original articles or reviews, and open access
status information derived from OurResearch's Unpaywall (Piwowar et al.,
2018), the same provider as OpenAlex. However, the Web of Science lacks
information about journals and articles under transformative agreements.

@@

Elsevier's Scopus, launched in 2004, is another widely used proprietary
bibliometric database for measuring research (Baas et al., 2020).
Similar to the Web of Science, Scopus is selective with regard to the
journals it indexes. However, its journal coverage is much broader than
that of the Web of Science collections considered in this study, as it
also indexes a wider range of applied research journals (Singh et al.,
2021; Stahlschmidt \& Stephen, 2022; Visser et al., 2021). With detailed
metadata about article types, open access status information derived
from Unpaywall, author roles, and disambiguated affiliations, Scopus
also contains important data to assess open access uptake, although no
direct information regarding transformative agreements was available at
the time of the study.

@@

Following Jahn (2025), the starting point was a unified dataset of
several safeguarded JCT snapshots\footnote{\url{https://github.com/njahn82/jct_data}},
as provided by hoaddata. The JCT journal data in hoaddata were enriched
with ISSN variants linked to an ISSN-L. To identify hybrid journals, a
comprehensive exclusion of fully open access journals was performed
using multiple journal lists (DOAJ, OpenAlex, Bielefeld). These data
were used to determine the publication volume of hybrid journals for
each database independently.

hoaddata relies on Crossref for obtaining journal publication volume and
open access status through Creative Commons licence information relative
to the published version (``version of record''). The article metadata
included DOIs, publication dates, open access information as well as
author roles and institutional affiliations. Publication years were
determined using the earliest known date of publication in a journal. In
hoaddata, this corresponded to Crossref's issued date. For Web of
Science and Scopus, the earliest publication date was used where
available, with Scopus dates specifically determined by the KB through
version tracking of the raw data.

Many transformative agreements typically cover only certain types of
journal articles, in particular original research articles including
reviews (Borrego et al., 2021). Because of limited information on these
document types in open scholarly data (Haupka et al., 2024), hoaddata
used an extended version of Unpaywall's paratext recognition approach to
exclude non-scholarly content (Piwowar et al., 2018). To exclude
conference supplements, which are also often not covered by
transformative agreements, only articles published in regular issues,
indicated by numerical pagination, were considered. For Web of Science
and Scopus, their established, mainly accurate document type
classifications (Donner, 2017; Maisano et al., 2025) were used to
identify original research articles and reviews, referred to as original
articles throughout this study.

@@

Articles in hybrid journals were considered open access when they were
made freely available under a Creative Commons license on publishers'
platforms. While hoaddata sourced this information from Crossref license
metadata, Web of Science and Scopus relied on Unpaywall as evidence
source. Unpaywall also uses Crossref license metadata, but supplements
them by parsing publisher websites directly, addressing cases where
publishers do not provide machine-readable Creative Commons license
information (Piwowar et al., 2018). This additional parsing remains
necessary despite transformative agreement workflows recommend the
depositing license information during DOI registration (Geschuhn \&
Stone, 2017). Both Web of Science and Scopus defined hybrid open access
consistently as content available under Creative Commons licenses on
publisher platforms according to their
documentations\footnote{\url{https://webofscience.help.clarivate.com/en-us/Content/open-access.html}}\footnote{\url{https://blog.scopus.com/posts/scopus-filters-for-open-access-type-and-green-oa-full-text-access-option}},
distinguishing it from bronze open access that lack such explicit
license information, or use publisher-specific licenses.

@@

Author affiliations were retrieved for both first and, if available,
corresponding authors to prepare the linking between articles and
institutions covered by transformative agreements. To improve the data
retrieval, JCT institution data was enriched with ROR-IDs from
associated institutions, such as university hospitals or institutes of
large research organisations such as the Max Planck Society, according
to OpenAlex' institution entity. To handle different address variants,
database-specific affiliation identifiers were used: ROR-IDs from
OpenAlex for hoaddata, affiliation enhanced names for Web of Science,
and Scopus Affiliation Identifier. Additionally, ISO country codes were
retrieved for each author's address to compile country-level statistics.
These country codes for Web of Science and Scopus were provided by the
KB.

@@

Based on these generated matching tables, articles eligible under
transformative agreements could also be retrieved from Web of Science
and Scopus, although they did not contain the ROR IDs used by the JCT.
The estimation of eligible articles followed Jahn (2025) and included a
matching of both journals and participating institutions. The matching
also took into account the duration of agreements according to the ESAC
registry, with only those matches where an agreement was actually in
place being considered for subsequent analysis. A related study (Jonge
et al., 2025), applied to publications funded by the Dutch Research
Council (NWO) and validated against internal invoicing data, confirmed
that such matching can accurately identify most articles under
transformative agreements.

@@

\centering
\begin{tabular}[t]{lrrr}
\toprule
\textbf{} & \textbf{hoaddata*} & \textbf{Web of Science} & \textbf{Scopus}\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Hybrid journal coverage}}\\
\hspace{1em}Active journals & 12,890 & 8,655 & 11,888\\
\hspace{1em}Active journals (core) & 12,888 & 8,655 & 11,878\\
\hspace{1em}Active journals (core) with OA & 11,348 & 8,392 & 11,313\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Publication volume}}\\
\hspace{1em}Total published articles & 9,740,015 & 8,616,053 & 8,117,644\\
\hspace{1em}Original articles & 8,158,425 & 6,708,083 & 7,317,703\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Digital Object Identifier (DOI) coverage}}\\
\hspace{1em}Articles with DOI & 9,740,015 & 7,713,796 & 8,105,112\\
\hspace{1em}Original articles with DOI & 8,158,425 & 6,695,661 & 7,314,327\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Open Access (OA) metrics}}\\
\hspace{1em}OA articles & 998,699 & 1,112,758 & 974,099\\
\hspace{1em}Original OA articles & 969,817 & 1,019,784 & 922,578\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Original articles with affiliation data}}\\
\hspace{1em}First author articles & 7,242,542 & 6,294,855 & 7,232,017\\
\hspace{1em}Corresponding author articles & 5,534,207 & 6,291,441 & 6,898,487\\
\bottomrule
\multicolumn{4}{l}{\rule{0pt}{1em}\textsuperscript{*} Journal article metadata from Crossref, except affiliations from OpenAlex}\\
\end{tabular}

@@

While hoaddata only covers articles with a DOI, Scopus and Web of
Science publication indicators were calculated using their database
identifier. A subsequent comparison of DOI coverage shows that
non-original articles in Web of Science often lacked a DOI. This was
particularly the case for meeting abstracts, which are notably prevalent
in Health Sciences journals (Melero-Fuentes et al., 2025) and are not
indexed by Scopus (Donner, 2017). Open access indicators were aggregated
by DOI, as Unpaywall only collects information on open access status for
articles with a DOI. A closer look at original articles with affiliation
data, and in line with related research (Zhang et al., 2024), reveals a
lack of affiliation data in the case of OpenAlex, the affiliation data
source used by hoaddata, compared to Web of Science and Scopus. In
particular, only about two-thirds of the articles examined provided
corresponding author affiliations. For first authors, the proportion was
89\%. At the time of writing, OpenAlex disclosed limited coverage of
corresponding authorhip data \footnote{\url{https://docs.openalex.org/api-entities/works/work-object/authorship-object\#is_corresponding}}.
Therefore, only first author data for hoaddata were considered in the
following analysis.

\subsection{Data analysis}\label{data-analysis}

@@

Journal coverage analysis revealed that 66\% (n = 7,970) of hybrid
journals included in transformative agreements were indexed in all three
databases (Figure \ref{fig-upset_coverage_results}A). The second-largest
set consisted of journals indexed in both hoaddata and Scopus,
comprising 21\% (n = 2,595) of hybrid journals. Notably, 6\% (n = 739)
of journals were exclusively contained in hoaddata, while another 6\% (n
= 748) were only found in the proprietory database Scopus. Of these, 354
were also available in the Web of Science. Upon inspection, this group
of hybrid journals exclusively covered in the proprietary data sources
mainly represented hybrid journals for which no open access evidence
could be retrieved from Crossref, the open access evidence source for
hoaddata.

@@

Using overlapping DOIs, the publication volume between 2019 and 2023 was
also calculated for each journal. Figure
\ref{fig-upset_coverage_results}C illustrates the distribution for each
combination. It shows that there is a large spread across the journals
covered by all three data sources. Furthermore, journals in this set
published more on average than journals covered by only one or two data
sources. In particular, the journals covered exclusively by hoaddata
were substantially smaller than those covered by all three sources. Upon
inspection, these were often newly launched hybrid journals, which
explains the relatively low five-year publication volume. An example is
\emph{Digital society} that published 86 articles. This hybrid journal
was launched in 2022, being covered by various Springer Nature
transformative agreements since then.

@@

Figure \ref{fig-upset_coverage_results_publisher} presents the coverage
of hybrid journals in transformative agreements across data sources from
2019 to 2023 with a focus on publisher portfolios. The analysis
highlights the dominance of the three largest publishers--Elsevier,
Springer Nature, and Wiley--, which collectively accounted for 47\% of
hybrid journals and 62\% of articles published during this five-year
period. In terms of article volume, Elsevier led with 2,441,358 articles
(33\% of the total) published across 1,951 hybrid journals (16\% of the
total). Springer Nature followed with 1,247,578 articles (17\%) in
2,311, although recording the largest number of hybrid journals (19\%).
Wiley accounted for 858,939 articles (12\%) in 1,382 hybrid journals
(11\%). The remaining 54 publishers collectively accounted for 2,749,847
articles (38\%) in 6,476 hybrid journals (53\%).

@@

Comparing publisher portfolios across different indexing sets
demonstrates that publisher were not represented uniformly. Notably,
Springer Nature exhibited 519 hybrid journals exclusively indexed in
both hoaddata and Scopus. This set included journals from the Chinese
Academy of Science, German-language medical journals, and Eastern
European publications including the \emph{Journal of Mathematical
Sciences}, which also publishes English-language translations of
Russian-language works. Additionally, this subset included titles with a
more broader disciplinary focus such as \emph{SN Computer Science} and
newly launched hybrid journals like \emph{Nature Computational Science},
which started in 2021 and was indexed in Scopus but not yet in Web of
Science. The set also captured ceased journals, providing further
insights into the dynamics of journal publishing.

@@

This section examines the uptake of open access in hybrid journals,
focusing on the influence of transformative agreements across hoaddata,
Scopus, and Web of Science. The aim was to assess whether consistent
results can be derived from these data sources despite differences in
coverage and methodologies. Following Jahn (2025), indicators were
calculated for each data source and comprise the number and proportion
of open access articles, including those enabled by transformative
agreements, from 2019 to 2023. For Web of Science and Scopus, the impact
of transformative agreements was estimated using both first and
corresponding authorships, while hoaddata indicator calculation was
limited to first author affiliations.

@@

Figure \ref{fig-uptake_overview}A shows a moderate growth of open access
in hybrid journals, which is consistent across hoaddata (black line),
Scopus (orange line), and Web of Science (blue line). According to
hoaddata, hybrid open access uptake increased from \% (n = ) in 2019 to
\% (n = ) in 2023. Similarly, Scopus recorded an growth from 6.5\% (n =
84,648) in 2019 to 19\% (n = 322,850) in 2023. However, Web of Science
recorded higher open access uptake in early years, before converging in
2023, from 11\% (n = 137,202) in 2019 to 20\% (n = 255,481) in 2023,
suggesting an different approach towards labeling hybrid open access by
the Web of Science.

Similarly, hybrid open access by transformative agreements increased
between 2019 and 2023 (Figures \ref{fig-uptake_overview}B and C). Trends
were consistent for first (dashed line) and corresponding author (dotted
line) affiliations. According to Scopus, 479,297 open access articles
could be attributed to transformative agreements based on first author
metadata (increasing from 15,341 to 206,084) and 489,262 using
corresponding author metadata (from 15,444 to 210,816). Web of Science
recorded 493,028 open access articles via transformative agreements
using first author metadata (increasing from 21,871 to 174,126) and
500,076 using corresponding author metadata (from 22,092 to 177,030).
hoaddata, lacking corresponding author data, linked 0 articles to
transformative agreements but showed slower growth than Scopus and Web
of Science, from 0 in 2019 to 0 in 2023.

From 2021 (hoadata, Scopus) resp. 2022 (Web of Science), transformative
agreements enabled the majority of hybrid open access. For first
authors, the share ranged between \% (hoaddata), 64\% (Scopus), and 68
\% (Web of Science) in 2023. For corresponding authors, the shares were
slightly larger, with Scopus recording 65\% and Web of Science 69\% in
2023. However, substantial hybrid open access was still facilitated
outside transformative agreements, likely through APCs paid from
discretionary research funds (Jahn et al., 2022; Suber, 2012).

@@

When considering open access trends by publisher (see Figure
\ref{fig-uptake_publisher}), the observed differences in early uptake
rates between hoaddata and Scopus compared to Web of Science can be
largely attributed to articles published in Elsevier hybrid journals,
the largest publisher in our sample. Both hoadata and Scopus reported an
steady increase in open access uptake between 2019 and 2023 (hoaddata
from 4\% to 13\%; Scopus from 5\% to 15\%). In contrast, Elsevier's
share remained relatively constant, increasing only slightly from 14\%
to 15\% according to the Web of Science. Upon inspection, this
discrepancy primarily stemmed from articles in the publisher's open
archive. These articles, made freely available after an embargo period
under Elsevier's user license, were tagged as hybrid open access in Web
of Science, even though its documentation\footnote{\url{https://webofscience.help.clarivate.com/en-us/Content/open-access.html}}
specified that only articles under a CC license variant were considered.
Previous research (Haustein et al., 2024; Jahn et al., 2022) has shown
that Elsevier provided a substantial portion of its articles under this
license, explaining the relatively large and stable share of open access
over the years.

@@

Despite these differences in open access evidence, the three data
sources show consistent temporal trends in hybrid open access enabled by
transformative agreements (see Figure \ref{fig-uptake_publisher}B and
C). Wiley emerged as the fastest-growing publisher in terms of open
access uptake, with more than 30\% of articles in hybrid journals
reported as open access in 2023 across the examined data sources,
followed by Springer Nature and Elsevier, recording a comparable late
uptake, which is consistent with the publisher's historical reluctance
to engage in negotations with library consortia. However, by 2023, the
share of open access enabled by transformative agreements appeared to
stabilise for all three publishers (see Figure
\ref{fig-uptake_publisher}C). Interestingly, the differences between
first and corresponding author affiliations were more pronounced at the
publisher level. In Scopus, for example, the share of open access via
transformative agreements measured by corresponding authorship was
greater for Springer Nature in 2023 than when using first authorship.

@@

When examining the percentage of open access articles in hybrid journals
(see Figure \ref{fig-uptake_country}B), a different pattern emerged.
Authors affiliated with institutions from medium-sized European
countries, such as Sweden, the Netherlands, Finland, and Hungary,
provided a large proportion of their articles in open access. Germany
and the United Kingdom also had approximately 40\% of their output
available as open access. In contrast, non-OECD countries showed notably
lower adoption of hybrid open access, with South Africa being the only
BRICS member well-represented in the data. The United States also
demonstrated a relatively low proportion of open access articles. These
findings were consistent across all databases. However, France was
better represented in Web of Science, likely due to its agreement with
Elsevier starting in 2019, which allowed delayed open access under the
publisher's user license (Rabesandratana, 2019). This licence was not
classified as hybrid open access in either Scopus or hoaddata. In all
cases, Spearman rank correlations were \(\rho > .9\), \(p < 0.001\),
showing a high level of correlations between the databases and
authorship roles considered.

Transformative agreements appeared to be a key driver of national open
access growth (see Figure \ref{fig-uptake_country}C and D). OECD members
accounted for the majority of open access articles enabled by
transformative agreements. As a notable exception, South Africa also
featured prominently, as the South African National Library and
Information Consortium (SANLiC) successfully negotiated transformative
agreements with major publishers from 2022 onward.\footnote{\url{https://sanlic.ac.za/read-and-publish-agreements/}}
Results were consistent across data sources. However, Wiley's open
access surplus in hoaddata led to better rankings for countries where
Wiley played a substantial role, such as Germany, where the DEAL
consortium negotiated its first transformative agreements with Wiley
that started in July 2019.

@@

\section{References}\label{references}

@@

\bibitem[\citeproctext]{ref-Borrego_2020}
Borrego, Á., Anglada, L., \& Abadal, E. (2021). Transformative
agreements: Do they pave the way to open access? \emph{Learned
Publishing}, \emph{34}(2), 216--232.
\url{https://doi.org/10.1002/leap.1347}

@@

@@

\bibitem[\citeproctext]{ref-de_Jonge_2025}
Jonge, H. de, Kramer, B., \& Sondervan, J. (2025). \emph{Tracking
transformative agreements through open metadata: Method and validation
using dutch research council NWO funded papers.}
\url{https://doi.org/10.31222/osf.io/tz6be_v1}

@@

\bibitem[\citeproctext]{ref-Maisano_2025}
Maisano, D. A., Mastrogiacomo, L., Ferrara, L., \& Franceschini, F.
(2025). A large-scale semi-automated approach for assessing
document-type classification errors in bibliometric databases.
\emph{Scientometrics}. \url{https://doi.org/10.1007/s11192-025-05244-y}

@@

\bibitem[\citeproctext]{ref-Stahlschmidt_2022}
Stahlschmidt, S., \& Stephen, D. (2022). From indexation policies
through citation networks to normalized citation impacts: Web of
science, scopus, and dimensions as varying resonance chambers.
\emph{Scientometrics}, \emph{127}(5), 2413--2431.
\url{https://doi.org/10.1007/s11192-022-04309-6}

\bibitem[\citeproctext]{ref-Suber_2012}
Suber, P. (2012). \emph{Open access}. The MIT Press.
\url{https://doi.org/10.7551/mitpress/9286.001.0001}
